---
title: "Graphical Discovery in Stochastic Actor-Oriented Models for Social Network Analysis"
author: "Sam Tyner"
date: "October 5, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
library(dplyr)
library(ggplot2)
library(geomnet)
library(network)
library(sna)
library(igraph)
library(tidyr)
library(RSiena)
library(knitr)
```

## About Me! 
* Bachelor of Arts, Augustana College, Rock Island, IL, 2012
    + Majors: Mathematics, Economics, French
* Master of Science, Iowa State University, Ames, IA, 2015
    + Major: Statistics
    + Committee: Heike Hofmann, Di Cook, Gray Calhoun (ECON)
* Passed PhD written examination, Summer 2014
* Aim to defend April 2017, graduate May 2017

## Thesis Outline

- Literature Review
- Chapter 1 - *Stochastic Actor-Oriented Models for Longitudinal Network Data: Removing the Blindfold*
- Chapter 2 - *Visual Inference for Stochastic Actor Oriented Models*
- Chapter 3 - *Drawing Networks in the `ggplot2` Framework*
- Other Projects 

## Outline of Talk

- Motivating Material
- Visual Inference for Stochastic Actor Oriented Models
- The `R` Package `geomnet`
- Thesis Timeline

# Motivating Material

## Social Networks

- "The Small World Problem" (Milgram 1967)
- Increased ease of data collection

## Statistical Models for Social Networks

- Erdös-Rényi Random Graph
- Exponential Random Graph
- Stochastic Actor-Oriented Models (SAOMs)

## Basic Notation

<div class = "columns-2">
Network: 
```{r, echo = FALSE, fig.align='center', out.width="95%", fig.height= 10}
set.seed(123456789)
nodes <- data.frame(id = LETTERS[1:5], x = c(1,1,2,2,3), y = c(1,2,1,2,1.5), label = 1:5)
edges <- data.frame(from = LETTERS[sample(5, 5)], to = LETTERS[sample(5, 5)])
net <- merge(edges, nodes, by.x = "from", by.y = "id", all = T)
edgelabels <- data.frame(label = LETTERS[1:5], 
                         x = c(.85,1.5,1.55,2.65,2.35), 
                         y = c(1.5, .9,1.45,1.8,1.65))
ggplot() + 
  geom_net(data = net, labelcolour = "forestgreen",
           layout = NULL, directed = T, size = 10, 
           color = 'black', shape = 1, vjust = .2, arrowgap = .08,
           ecolour = 'grey40', curvature = .1, 
           aes(x = x, y = y, from_id = from, to_id = to, label = label)) + 
  #geom_text(data = edgelabels, size = 5,
  #          aes(x = x, y = y, label = label)) + 
  theme_net() + theme(plot.title = element_text(size = 40)) + 
  labs(title = "x")
```

- Nodeset: $\mathcal{N} = \{1,2,3,4,5\}$
- Edgeset: $\mathcal{E} = \{x_{ij}\}$ for $i \neq j \in \mathcal{N}$
- Binary, directed edges: $x_{12} = 1 \neq x_{21} = 0$ 
- Observed network: $x$
- Unobserved network (random variable): $X$
</div>

## Erdös-Rényi
For undirected graphs 

- Start with $n$ nodes
- $x_{ij}$ chosen at random from $\binom{n}{2}$ possible
- $Pr(x_{ij} = 1) = p$ for any $i \neq j \in \{1, \dots, n\}$
- $e = \sum_{i < j} x_{ij}$, the number of edges has likelihood:

$$ f(e| p, n) = p^{e}(1-p)^{\binom{n}{2} - e} $$



# Visual Inference for SAOMs

# The `R` package `geomnet`

# Closing Material
