---
title: "Dissertation Proposal"
author: "Samantha C. Tyner"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_book:
    keep_tex: no 
    toc: true
    number_sections: true
    includes:
      before_body: body.tex
header-includes:
- \usepackage{booktabs}  
bibliography: docs/bigbib.bib

---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
library(dplyr)
library(ggplot2)
library(geomnet)
library(tidyr)
library(RSiena)
library(knitr)
```
# Outline of the whole shebang
This section will be removed once I'm done writing this thing! 

## Lit Review (CH. 0)

### ~~High Level Introduction~~

~~Social networks have been studied for decades, beginning with the seminal 1967 study, "The Small World Problem" by Stanley Milgram  (Goldenberg et al. 2009). In recent years, the study of social networks has grown in popularity due to an increase in the availability of and easy of access to social network data. ~~

* ~~discuss the data format and how it is different from traditional data formats seen in statistics~~
* ~~bring up the different disciplines that have studied social networks~~
* ~~mention the models that exist for social network analysis~~
* ~~mention that I am focusing on one type of model class, SAOMs.~~

### More Meaty Introduction to Network Analysis

In this section I will discuss in greater detail the models that are out there for network analysis. These include:

* ~~Classics:~~
    + ~~ERGMs~~
    + ~~Erdos-Renyi~~ and its variations
    + ~~Random graphs with fixed degree distribution~~
    + ~~Blockmodels and their relatives~~
    + ~~Latent space models~~
* ~~Social Networks:~~
    + ~~"$p_1$" models~~
    + ~~"$p_2$" models~~
* Dynamic:
    + ~~Preferential attachment~~
    + ~~Small-world~~
    + ~~Duplication-attachment~~
    + Discrete time MC models
    + Continuous time MC models (SAOMs a kind of CTMC)
  
### Why focus on SAOMs

First, I will layout the field of network analysis formally using traditional statistical paradigms, as discussed in Kolaczyk (2009). Then I will discuss the lack of model checking tools specific to SAOMs due to model intractibility. This will be brief and will lead into the next section, a complete introduction to SAOMs and their structure/properties. 


### Full introduction and discussion of SAOMs

I will completely flesh out the form and theory of SAOMs. This will come primarily from the detailed document I created last year and anything else from Snijders' papers. 

### Lead-in to Visual Inference 

Take the theory of SAOMs and lead into the discussion of why visual inference could be useful for model checking and other things for SAOMs. 

### Introduction to Visual Inference

Fully introduce the concept of visual inference as laid out in Buja et al. (2009). Also include discussion of Majumder et al. 

### Intro to visualizing networks

Discuss the many ways to visualize networks (also called network mapping). Ubiquitous view is 2D represetation of points and lines. Discuss the many layout algorithms. Discuss the many packages in R that can do this. Discuss the difficulty of using these package for people not intimately familiar with them. Lead into next section with discussion of ggplot2 and its importance and popularity in data visualization.

### Lead-in to the full thesis. 

Briefly outline the 3 chapters and tie them all together. 

## CH 1. Removing the Blindfold from SAOMs

- What does the distribution look like? 
- What does an "average network" from an SAOM look like? 
- Generation / Creation process video

## CH 2. Using visual inference for SAOMs

Explain how I will use / have used visual inference to explore properties of SAOMs. Main research question: are statistically significant model differences also visually significant? 


## CH. 3: geomnet: a ggplot2 wrapper for network visualization

Explain the hole filled by the geomnet package. 

## CH. 4 Summary, Impact, Future Work

Brief walk-through of all topics already covered. Emphasize the holes filled by the work. 

## CH 5. Other Projects  

Discuss the Bob Ross and Graphics Awards papers


\newpage

```{r child = "docs/litreview.Rmd", eval=TRUE}
```

```{r child = "docs/removingblindfoldsaoms.Rmd", eval = TRUE}
```

```{r child = "docs/visualinferencenets.Rmd", eval = TRUE}
```

```{r child= "../../NetworksResearch/ggnet-paper/RJwrapper.pdf", eval = FALSE}
# change options above until you figure out how to get the stupid pdf in there. 
```

```{r child = "docs/futurework-and-otherwork.Rmd", eval=TRUE}
```

# References {#refs}

