---
title: "Graphical Discovery in Stochastic Actor-Oriented Models for Social Network Analysis"
author: "Samantha C. Tyner"
date: "`r Sys.Date()`"
knit: "bookdown::render_book"
output:
  bookdown::pdf_book:
    keep_tex: no
    toc: true
    number_sections: true
    includes:
      before_body: body.tex
header-includes:
- \usepackage{booktabs} 
- \usepackage{pdfpages} 
- \usepackage{graphicx}
- \usepackage{epstopdf}
bibliography: docs/bigbib.bib
link-citations: true

---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
library(dplyr)
library(ggplot2)
library(geomnet)
library(network)
library(sna)
library(igraph)
library(tidyr)
library(RSiena)
library(knitr)
```

\newpage

```{r child = "docs/litreview.Rmd", eval=TRUE}
```

\newpage 

```{r child = "../SAOM-removing-blindfold/paper-child.Rnw", eval = TRUE}
```

\newpage

```{r child = "docs/visualinferencenets.Rmd", eval = TRUE}
```

# Drawing Networks in the `ggplot2` Framework {#ch3}

This next section is a paper I authored with Heike Hofmann (Iowa State University) and François Briatte (European School of Political and Social Sciences) that has been accepted for publication in *The R Journal* subject to revision. I developed the package `geomnet` ("our package") with Dr. Hofmann in 2015. Separately from the implementation of the graph `geom` in our package, Briatte implemented graph visualization in two different approaches: the `ggnet2` function in the `GGally` package and the `ggnetwork` package. This paper gives an overview of the three different approaches, highlights why `ggplot2` is generally well suited for graph visualization, and then discusses the pros and cons of the three methods: `geom_net()` in `geomnet`, `ggnet2()` in `GGally`, and `ggnetwork()` in `ggnetwork`. 

\newpage

# Other Projects

I have also completed a couple of other significant research projects that I feel have been important in my studies. The first of these is a project on clustering digital images of paintings of the artist Bob Ross. This was originally a project assignment for STAT 503: Exploratory Methods and Data Mining in Spring 2015. I entered a version of the project assignment into Significance Magazine's 2015 Young Statisticians Writing Competition, where it was among the final three choices for the top prize.^[See [https://www.statslife.org.uk/culture/2553-the-joy-of-clustering](https://www.statslife.org.uk/culture/2553-the-joy-of-clustering).] The other project is a visual exploration of the Trans-Atlantic Slave Trade Database, in which I used the `geomnet` package to better understand the structure and impact of the trans-atlantic slave trade. I submitted this paper to the 2016 Student Paper Competition sponsored by the ASA Section on Statistical Graphics and Computing, where it won top prize for a student paper in Graphics.^[See [http://stat-computing.org/awards/student/winners.html](http://stat-computing.org/awards/student/winners.html).]  

# References {#refs}

