---
title: "Graphical discovery in stochastic actor-oriented models for social network analysis"
author: "Sam Tyner"
date: "2017/11/07"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: samdefense.css  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, message = FALSE, warning = FALSE, dev = 'png', fig.width = 14, fig.height=9, fig.align='center')
```

```{r packages}
library(tidyverse)
library(geomnet)
library(RSiena)
library(lme4)
library(vinference)
```

```{r loaddata}
load("data/finalglmm31-2.RDA")
mod <- model2randomscalesize
turk22 <- read_csv("data/turk22-sig.csv")
```

# Updates

Publications: 

- **Tyner, S.**, Briatte, F. and Hofmann, H. "Network Visualization with ggplot2." *The R Journal*. 9:1, 27-54, URL: https://journal.r-project.org/archive/2017/RJ-2017-023/index.html  
--

- **Sam Tyner** (2016) "Using the R Package geomnet: Visualizing Trans-Atlantic
Slave Trade of Africans, 1514-1866," *CHANCE*, 29:3, 4-16, DOI: 10.1080/09332480.2016.1234879

---

# Updates

Post-graduation plans:

- Become a post-doctoral researcher in CSAFE at Iowa State working on human factors: 
    * Develop training materials for judges, lawyers, & practitioners 
    * Design and conduct mock jury studies
    * Open source, reproducible research in forensics
    
---

# Outline 

- Remaining results from MTurk Experiment
- Shiny app for creating lineups
- Shiny app for visual inference experiment
- Model Visualization for SAOMs
    * Model-in-data-space vs. Data-in-model-space
    * Collections of models
    * Exploring algorithms 
- Developing animations: *netvizinf* package 

---
class: inverse, center, middle

# Results (continued)

---

# Significance tests

- $p$-values calculated using the *vinference* package (Hofmann and Röttger, 2016)

```{r sigres, results='asis'}
turk22_sig <- turk22 %>% 
  mutate(type2 = ifelse(sign == 0, sign(initialEst), sign),
         type2 = ifelse(type == "one", type2, -1)) %>% 
  filter(param_value == 1)
summ_sig <- turk22_sig %>% group_by(pic_id, test_param) %>% 
  summarize(npick = sum(datapick), n = n(), 
            pv = map2_dbl(npick, n, pV, m=6, scenario=3))
names(summ_sig) <- c("Lineup ID", "parameter", "# Alt. Model Picks", "Total Views", "p-value")
summ_sig$parameter <- as.factor(summ_sig$parameter)
levels(summ_sig$parameter) <- paste0("&beta;<sub>", c(3,4), "</sub>")
knitr::kable(summ_sig, format = "markdown", digits = c(0,0,0,0,5))
```

---
class: plotslide

# Significance tests - $\beta_3$, rep. 1

- 4/29 chose alternative data plot (#2)

```{r sigjttp1}
dat <- read_csv("data/jttp_pos_hard_1.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Significance tests - $\beta_3$, rep. 2

- 26/31 chose alternative data plot (#3)

```{r sigjttp2}
dat <- read_csv("data/jttp_pos_hard_2.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Significance tests - $\beta_3$, rep. 3

- 2/27 chose alternative data plot (#1)

```{r sigjttp3}
dat <- read_csv("data/jttp_pos_hard_3.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Significance tests - $\beta_4$, rep. 1

- 10/23 chose alternative data plot (#5)

```{r sigjtts1}
dat <- read_csv("data/jtts_pos_hard_1.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Significance tests - $\beta_4$, rep. 2

- 3/37 chose alternative data plot (#1)

```{r sigjtts2}
dat <- read_csv("data/jtts_pos_hard_2.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Significance tests - $\beta_4$, rep. 3

- 10/29 chose alternative data plot (#2)

```{r sigjtts3}
dat <- read_csv("data/jtts_pos_hard_3.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---

# Goodness-of-fit tests

- $p$-values calculated using the *vinference* package (Hofmann and Röttger, 2016)

```{r gof, results='asis'}
turk22_gof <- read_csv("data/turk22-gof.csv")
turk22_gof_stats <- turk22_gof %>% group_by(pic_id) %>% 
  summarize(npickdata = sum(datapick), total = n(), 
          pvinf = map2_dbl(npickdata, total, pV, m = 6, scenario = 3))
turk22_gof_stats %>% separate(pic_id, into = c("discard", "model", "rep"), sep = c(2,3)) -> print_gof_stats
print_gof_stats %>% select(-discard) %>% mutate(model = paste0("M", (as.integer(model))))-> print_gof_stats
names(print_gof_stats) <- c("Model", "Replicate", "Data Picks", "Total Viewers", "p-value")
print_gof_stats$`p-value` <- ifelse(print_gof_stats$`p-value` < .0001, "< 0.0001", as.character(round(print_gof_stats$`p-value`,4)))
knitr::kable(print_gof_stats, format="markdown", digits = 4, align = c("l", "c", "c", "c", "r"))
```

---
class: plotslide

# Goodness-of-fit test: M3, rep 1

- 29/36 selected the data

```{r gofm31}
load("data/se112adjmat.RDS")
wave2 <- data.frame(se112adj)
senators <- data.frame(id = colnames(wave2), number = 1:155)
wave2$from <- colnames(wave2)
wave2 <- wave2 %>% gather(to, val, Alan.Stuart.Franken:William.Cowan) %>% filter(val > 0)
wave2 <- merge(wave2, senators, by.x = "from", by.y = "id", all = T)
wave2$from <- senators$number[match(wave2$from, senators$id)]
wave2$to <- senators$number[match(wave2$to, senators$id)]
wave2$from <- paste0("V", wave2$from)
wave2$to <- ifelse(is.na(wave2$to), NA, paste0("V", wave2$to))
wave2 <- wave2 %>% mutate(sim = 1001, model = "data", wave = 1) %>% select(-c(val, number))
dat <- read_csv(paste0("data/jttp_gof_9_1.csv"))
datplot <- unique(dat$ord[dat$sim == 1001])
dat <- dat %>% filter(sim != 1001) %>% 
      bind_rows(wave2) %>% 
      mutate(ord = ifelse(is.na(ord), datplot, ord))
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Goodness-of-fit test: M4, rep 2

- 7/20 selected the data 

```{r}
dat <- read_csv(paste0("data/jtts_gof_9_2.csv"))
datplot <- unique(dat$ord[dat$sim == 1001])
dat <- dat %>% filter(sim != 1001) %>% 
      bind_rows(wave2) %>% 
      mutate(ord = ifelse(is.na(ord), datplot, ord))
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Goodness-of-fit test: M5, rep 3

- 14/16 selected the data 

```{r}
dat <- read_csv(paste0("data/simttb_gof_9_3.csv"))
datplot <- unique(dat$ord[dat$sim == 1001])
dat <- dat %>% filter(sim != 1001) %>% 
      bind_rows(wave2) %>% 
      mutate(ord = ifelse(is.na(ord), datplot, ord))
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: plotslide

# Goodness-of-fit test: M7, rep 1

- 17/20 selected the data 

```{r}
dat <- read_csv(paste0("data/bigmod_gof_9_1.csv"))
datplot <- unique(dat$ord[dat$sim == 1001])
dat <- dat %>% filter(sim != 1001) %>% 
      bind_rows(wave2) %>% 
      mutate(ord = ifelse(is.na(ord), datplot, ord))
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size =1) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black'),
              strip.text = element_text(size = 20)) +
        facet_wrap(~ord)
```

---
class: inverse, center, middle

# App for creating lineups

---
class: inverse, center, middle

# Shiny app for visual inference experiment

---
class: inverse, center, middle

# Model Visualization for SAOMs

---
class: definition

# Model visualization 

## Model visualization summarizes the model with graphical, as opposed to numerical summaries to answer questions such as

1. What does the model look like? 
2. How does the model change when its parameters change? How do the parameters change when the data is changed?
3. How well does the model fit the data? How does the shape of the model compare to the shape of the data? 
4. Is the model fitting uniformly good, or good in some regions but poor in other regions? Where might the fit be improved?

(questions from Wickham et al 2015)

---

# How? 

Three main approaches: 

- View the model in the data-space
- View collections of models
- Exploring algorithms 


---

# M-in-DS

- Model-in-data-space vs. Data-in-model-space

---

# Developing animations: *netvizinf* package 


---

# References

Hofmann, H. and Röttger, C. (2016). vinference: Inference under the
  lineup protocol. R package version 0.1.1. http://github.com/heike/vinference

